### 2016-11-15 greedy_hamiltonian_revised/chapters/4.sympmor.tex
fist commit

\begin{proof}
	Let $\phi_t$ and $\psi_t$ be the Hamiltonian flow of the original and the reduced system respectively. By definition $\mathbf z(t) = \phi_t(\mathbf z_0)$ and $\mathbf y(t) = \psi_t(\mathbf y_0)$. Using Theorem \ref{theorem:Hasy:1} we have
\begin{equation} \label{eq:SyMo:6}
\begin{aligned}
	H(\tilde{\mathbf{z}} (t)) = H( A\mathbf y (t) ) = \tilde H(\mathbf y (t)) = \tilde H(\psi_t(\mathbf y_0)) = \tilde H(\mathbf y_0) = \tilde H(A^+ \mathbf z_0) = H(AA^+\mathbf z_0).
\end{aligned}
\end{equation}
Also $H(\phi_t(\mathbf z_0)) = H(\mathbf z_0)$. The error in the Hamiltonian can be written in terms of $\mathbf z_0$ and the symplectic basis $A$ as
\begin{equation} \label{eq:SyMo:7}
	\Delta H(t) = |H(\mathbf z_0) - H(AA^+\mathbf z_0)|
\end{equation}
\end{proof}

The following theorems guarantees stability of the original and the reduced Hamiltonian systems. We refer the reader to \cite{Peng:2014di} for the proofs.

\begin{theorem} \label{theorem:SyMo:1}
Consider the initial value problem (\ref{eq:Hasy:7}) with its symplectic reduced system (\ref{eq:SyMo:4}). If we can find a bounded neighborhood $N$ of $z_0$ in $P$ such that $H(z) < H(z_0)$ or $H(z) > H(z_0)$, for all $z\in \partial N$, then the original system and the reduced system are bounded for all $t\in \mathbb R$.
\end{theorem}

\begin{theorem} \label{theorem:SyMo:2}
If the original system (\ref{eq:Hasy:7}) is a linear Hamiltonian system, i.e. $H(x) = \frac 1 2 x^T L x$ with $L$ a $2n\times 2n$ real symmetric matrix, then both the original system and the reduced system (\ref{eq:SyMo:4}) are stable.
\end{theorem}

@@

Similar to Section \ref{chap:MoOr.PrOr:1} we gather snapshots $\mathbf z_i = [q_i^T , p_i^T]^T$ in the snapshot matrix $S$. Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided. The proper symplectic decomposition requires that the error of the symplectic projection onto the symplectic subspace to be minimum. Analogous to the minimization (\ref{eq:MoOr:6}), the PSD symplectic basis of size $2k$ is the solution of the optimization problem

\begin{equation} \label{eq:SyMo:8}
\begin{aligned}
& \underset{V\in \mathbb R^{2n\times 2k}}{\text{minimize}}
& & \| S - AA^+S\|_F \\
& \text{subject to}
& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}
\end{aligned}
\end{equation}
Note $A$ is not necessarily an orthogonal matrix. Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$. At first, the minimization looks similar to the one obtained by POD. However, the lack of an SVD-like decomposition for symplectic subspaces, makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).

As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, direct solution usually expensive. A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis. In the following we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$. Finally, we introduce a greedy approach for symplectic basis generation.

@@

%Algorithm \ref{alg:SyMo:1} summarizes the cotengent lift method for finding a symplectic basis.
%
%\begin{algorithm} 
%\caption{Real SVD method for symplectic basis generation} \label{alg:SyMo:1}
%{\bf Input:} Snapshots $\{ q_i , p_i \}_{i=1}^N$ of the solution manifold and $2k$ the size of the symplectic basis.
%\begin{enumerate}
%\item Construct the combined snapshot matrix $S_{\text{combined}}$ in (\ref{eq:SyMo:10}).
%\item Compute the SVD decomposition of $S_{\text{combined}}=U\Sigma V^T$.
%\item Define $\Phi = [u_1,\dots,u_k]$ with $u_i$ the $i$-th left singular vector.
%\item Construct $A$ according to (\ref{eq:SyMo:9}).
%\end{enumerate}
%\vspace{0.5cm}
%{\bf Output:} Symplectic matrix $A$ of the size $2n\times 2k$.
%\end{algorithm}
%

\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form
\begin{equation} \label{eq:SyMo:11}
	A = 
	\begin{pmatrix}
		\Phi & -\Psi \\
		\Psi & \Phi
	\end{pmatrix},
\end{equation}
such that $\Phi$ and $\Psi$ are real matrices of the size $n\times k$ satisfying conditions
\begin{equation} \label{eq:SyMo:12}
\Phi^T \Phi + \Psi^T \Psi = I_k,\quad \Phi^T \Psi = \Psi^T \Phi.
\end{equation}
It can be checked that $A$ forms a symplectic matrix. To construct $A$ we first define the complex snapshot matrix
\begin{equation} \label{eq:SyMo:13}
	S_{\text{complex}} = [ q_1 + i p_1, \dots , q_N + i p_N ].
\end{equation}
The SVD decomposition of $S_{\text{complex}}$ can be written as $S_{\text{complex}} = U \Sigma V^H$ with the superscript $H$ indicating the conjugate transpose of a complex valued matrix. Each left singular vector of $S_{\text{complex}}$ takes the form $u_m = r_m + i s_m$. We define
\begin{equation} \label{eq:SyMo:14}
	 \Phi = [r_1,\dots, r_k], \quad \Psi = [s_1,\dots, s_k].
\end{equation}
Then we can easily check that conditions (\ref{eq:SyMo:12}) is satisfied since $U$ is unitary. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.

%Algorithm \ref{alg:SyMo:2} summarizes the complex SVD method for finding a symplectic basis.
%
%\begin{algorithm} 
%\caption{Complex SVD method for symplectic basis generation} \label{alg:SyMo:2}
%{\bf Input:} Snapshots $\{ q_i , p_i \}_{i=1}^N$ of the solution manifold and $2k$ the size of the symplectic basis.
%\begin{enumerate}
%\item Construct the complex snapshot matrix $S_{\text{complex}}$ in (\ref{eq:SyMo:13}).
%\item Compute the complex SVD decomposition $S_{\text{complex}} = U \Sigma V^H$.
%\item Define $\Phi$ and $\Psi$ as in (\ref{eq:SyMo:14}).
%\item Construct $A$ according to (\ref{eq:SyMo:11}).
%\end{enumerate}
%\vspace{0.5cm}
%{\bf Output:} Symplectic matrix $A$ of the size $2n\times 2k$.
%\end{algorithm}

\subsubsection{The Greedy Approach to Symplectic Basis Generation} The greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This results in more compact bases and/or more accurate reduced systems. In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.

{\bf Symplectic Gram-Schmidt:} As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}). Suppose that $S=\{ e_1 , \dots , e_k \} \bigcup \{ f_1 , \dots , f_k \}$ is a symplectic basis and $a$ and $b$ are two vectors such that $a,b \not \in \text{span}(S)$ and $a \not \not \parallel b$. We seek to grow the size of $S$ by the vectors $a$ and $b$. First we modify $a$ and $b$ to ensure they are symplectic (symplectic orthogonal) to $S$, i.e., we add a linear combination of elements of $S$ to $a$ and $b$, so the resulting vectors are symplectic to $S$. This can be achieved since we assumed $a,b\not \in \text{span}(S)$. So for $a$ we seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation}
	\Omega\left(a + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i f_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
\begin{equation}
	\alpha_i = - \Omega(a,f_i), \quad \beta_i = \Omega(a,e_i),
\end{equation}
for $i=1,\dots,k$. Now we define the modified vectors as
\begin{equation}
\begin{aligned}
	& \tilde a = a - \sum_{i=1}^k \Omega(a,f_i) e_i + \sum_{i=1}^k \Omega(a,e_i) f_i, \\
	& \tilde b = b - \sum_{i=1}^k \Omega(b,f_i) e_i + \sum_{i=1}^k \Omega(b,e_i) f_i.
\end{aligned}
\end{equation}
Finally we construct the new basis vectors as
\begin{equation}
\begin{aligned}
	& e_{k+1} = \text{sign}(\Omega(\tilde a , \tilde b)) \cdot \frac{\tilde a}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }, \quad f_{k+1} = \frac{\tilde b}{ \sqrt{ |\Omega(\tilde a , \tilde b)| } }.
\end{aligned}
\end{equation}
One easily checks that the new basis vectors $e_1,\dots,e_{k+1},f_1,\dots,f_{k+1}$ satisfy all the conditions (\ref{eq:Hasy:4}). Similar to the QR decomposition, the symplectic Gram-Schmidt process leads to a matrix decomposition for any full ranked matrix of size $2n\times 2k$. More details about this matrix decomposition are given in Appendix \ref{chap:Append:1}.

In the case of finite symplectic vector fields, there is an alternative way to update the symplectic basis $S$. To illustrate, we first note that if $e_i$ or $f_i$ is a basis vector in $S$, we have $\mathbb J_{2n}e_i = -f_i$ and $\mathbb J_{2n}f_i = e_i$. Since
\begin{equation}
	e_i^T\mathbb J_{2n} f_i = \Omega(e_i,f_i) = 1 \Longleftrightarrow \mathbb J_{2n} e_i = - f_i \text{ and } \mathbb J_{2n} f_i = e_i.
\end{equation}
Thus if we consider a vector $v\not \in \text{span}(S)$ then $\mathbb J_{2n}^T v $ also does not belong to span$(S)$. Since, if it did, we could write the vector as a linear combination of basis vectors of $S$ as
\begin{equation}
	\mathbb J_{2n}^T v = \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i f_i.
\end{equation}
Multiplying both sides by $\mathbb J_{2n}$ yields
\begin{equation}
	v = \sum_{i=1}^m \alpha_i \mathbb J_{2n} e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n} f_i,
\end{equation}
But all vectors on the right hand side belong to span$(S)$, hence reaching a contradiction. For updating $S$ using $v$, we first symplectify (symplectic orthogonalize) $v$ with respect to $S$ to recover
\begin{equation}
	\tilde v = v - \sum_{i=1}^k \Omega(v,f_i)e_i + \sum_{i=1}^k \Omega (v,e_i)f_i.
\end{equation}
We can write the update for the symplectic basis as
\begin{equation}
	e_{k+1} = \frac{\tilde v}{\|\tilde v\|},\quad f_{k+1} = \mathbb J_{2n}^T \tilde v.
\end{equation}
It is easily checked that the new set of basis vectors $\{e_1,\dots,e_{k+1},f_1,\dots,f_{k+1} \}$ forms a symplectic basis. This update, not only provides a symplectic basis but also an orthonormal basis.

@@

Suppose that a $2k$ dimensional symplectic basis $A_{2k} = \{e_1,\dots ,e_k,f_1,\dots ,f_k\}$ is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors. Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to find the value that maximises the error in the Hamiltonian
\begin{equation}
	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
\end{equation}
The goal is to approximate the Hamiltonian function as well as possible.

We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots $S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}$. The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
\begin{equation}
	z := \underset{s\in S}{\text{argmax }} \| s - AA^+s \|.
\end{equation}
Finally, we update the basis as
\begin{equation}
	e_{k+1} = \tilde z, \quad f_{k+1} = \mathbb J_{2n}^T \tilde z,
\end{equation}
where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.

@@

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Gamma_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $f_1 \leftarrow \mathbb J^T_{2n} \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,f_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
\item \hspace{0.5cm}$f_{k+1} \leftarrow \mathbb J^T_{2n} \mathbf z^*$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , f_1,\dots,f_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}

The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}. We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
\begin{itemize}
\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases. Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis. The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
\end{itemize}

@@

	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $d_{n}\leq c\exp(-\alpha k)$ with $\alpha > \log3$. Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provide exponential approximation properties such that

@@

>>>>

\begin{proof}
	Let $\phi_t$ and $\psi_t$ be the Hamiltonian flow of the original and the reduced system respectively. By definition $\mathbf z(t) = \phi_t(\mathbf z_0)$ and $\mathbf y(t) = \psi_t(\mathbf y_0)$. Using the definition of the reduced Hamiltonian and Theorem \ref{theorem:Hasy:1} we have
\begin{equation} \label{eq:SyMo:6}
\begin{aligned}
	H(\tilde{\mathbf{z}} (t)) = H( A\mathbf y (t) ) = \tilde H(\mathbf y (t)) = \tilde H(\psi_t(\mathbf y_0)) = \tilde H(\mathbf y_0) = \tilde H(A^+ \mathbf z_0) = H(AA^+\mathbf z_0).
\end{aligned}
\end{equation}
The error in the Hamiltonian can then be written in terms of $\mathbf z_0$ and the symplectic basis $A$ as
\begin{equation} \label{eq:SyMo:7}
	\Delta H(t) = |H(\mathbf z_0) - H(AA^+\mathbf z_0)|
\end{equation}
\end{proof}

{\edit 
The following theorems provide a strong indication of the stability of the reduced system.

\begin{definition} \label{definition:SyMo:1}
Consider a dynamical system of the form $\dot{\mathbf z} = \mathbf f(\mathbf z)$ and suppose that $\mathbf z_e$ is an equilibrium point for the system so that $\mathbf f(\mathbf z_e) = 0$. $\mathbf z_e$ is called nonlinear stable or Lyapunov stable if for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $t\geq0$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
\end{definition}	
The following proposition, also known as Dirichlet's theorem \cite{bhatia2002stability}, states the sufficient condition for an equilibrium point to be Lyapunov stable. We refer the reader to \cite{bhatia2002stability} for the proof.
\begin{proposition} \label{proposition:SyMo:1}
An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exist a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, and $\nabla^2 W(\mathbf z_e)$ is positive definite and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$. Here $\nabla^2W$ is the Hessian matrix of $W$.
\end{proposition}
The scalar function $W$ is referred to as the \emph{Lyapunov function}. In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$. The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Liyapunov stable.
\begin{theorem} \label{theorem:SyMo:1}
Consider a Hamiltonian system of the form (\ref{eq:Hasy:7}) together with the reduced system (\ref{eq:SyMo:4}). Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$. If $H$ (or $-H$) is a Lyapunov function satisfying proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}) respectively.
\end{theorem}
\begin{proof}
	It is the direct consequence of proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point. It can be easily checked that if $\mathbf z_e$ is a local minimum of $H$ then $\mathbf y_e$ is a local minimum for $\tilde H$ and an equilibrium point for (\ref{eq:SyMo:4}). Also from the chain rule we have
\[
	\nabla^2_{\mathbf y} \tilde H = A^T \nabla^2_{\mathbf z} H A.
\]
So for any $\xi\in \mathbb R^{2k}$
\[
	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0,
\]
Where the last inequality is due to the positive definiteness of $H$. Therefore $\tilde H$ is also positive definite. By proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
\end{proof}
}

@@

Similar to Section \ref{chap:MoOr.PrOr:1} we gather snapshots $\mathbf z_i = [q_i^T , p_i^T]^T$ in the snapshot matrix $S$. Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided. {\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace to be minimum. Analogous to the minimization (\ref{eq:MoOr:6}), the PSD symplectic basis of size $2k$ is the solution of the optimization problem

\begin{equation} \label{eq:SyMo:8}
\begin{aligned}
& \underset{V\in \mathbb R^{2n\times 2k}}{\text{minimize}}
& & \| S - AA^+S\|_F \\
& \text{subject to}
& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}
\end{aligned}
\end{equation}
Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$. At first, the minimization looks similar to the one obtained by POD. {\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded, this means numerical errors can become dominant in symplectic projection \cite{Karow:2006cf}. This makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}

As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, direct solution usually expensive. A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis.

{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices is proposed by \cite{Xu:2003kx}. However, computation of a possibly large snapshot matrix and then using the mentioned methods to compute its eigen-spaces, is usually computationally demanding. Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.

The greedy approach presented in section \ref{Chap:Symo.PrSy:3}, is an iterative method for construction of a symplectic basis. It avoids evaluation of the full snapshot matrix so it saves substantial computational cost in the offline stage of the symplectic model reduction. Also, by construction, it yields a orthosymplectic basis and therefore a well-conditioned basis.

In section \ref{chap:SyMo.PrSy:2} we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$. Finally, in section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for symplectic basis generation.}

@@

\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form \cite{Peng:2014di}
\begin{equation} \label{eq:SyMo:11}
	A = 
	\begin{pmatrix}
		\Phi & -\Psi \\
		\Psi & \Phi
	\end{pmatrix},
\end{equation}
such that $\Phi$ and $\Psi$ are real matrices of the size $n\times k$ satisfying conditions
\begin{equation} \label{eq:SyMo:12}
\Phi^T \Phi + \Psi^T \Psi = I_k,\quad \Phi^T \Psi = \Psi^T \Phi.
\end{equation}
It can be checked that $A$ forms a symplectic matrix. To construct $A$ we first define the complex snapshot matrix
\begin{equation} \label{eq:SyMo:13}
	S_{\text{complex}} = [ q_1 + i p_1, \dots , q_N + i p_N ].
\end{equation}
The SVD decomposition of $S_{\text{complex}}$ can be written as $S_{\text{complex}} = U \Sigma V^H$ with the superscript $H$ indicating the conjugate transpose of a complex valued matrix. Each left singular vector of $S_{\text{complex}}$ takes the form $u_m = r_m + i s_m$. We define
\begin{equation} \label{eq:SyMo:14}
	 \Phi = [r_1,\dots, r_k], \quad \Psi = [s_1,\dots, s_k].
\end{equation}
Then we can easily check that conditions (\ref{eq:SyMo:12}) is satisfied since $U$ is unitary. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.

\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This results in more compact bases and/or more accurate reduced systems. In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.

{\edit The orthonormalization step is an essential step in most greedy approaches for basis generation in the context of model reduction \cite{Anonymous:2016wl,Quarteroni:2016wi}. However common orthonormalization processes, e.g. the QR method based on the classical Gram-Schmidt method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}. Here we use a variation of the QR method known as the the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis. 
}

{\edit
As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}). Further theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}. To briefly describe the SR method, suppose that an orthosymplectic basis
\begin{equation} \label{eq:SyMo:14.1}
	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ -\mathbb J_{2n} e_1 , \dots , -\mathbb J_{2n} e_k \},
\end{equation}
and a vector $z\not \in \text{span}(A_{2k})$ is provided. We would like to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$. We seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation} \label{eq:SyMo:14.2}
	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i - \sum_{i=1}^k \beta_i \mathbb J_{2n}e_i , \sum_{i=1}^k \bar{\alpha}_i e_i - \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}e_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
\begin{equation} \label{eq:SyMo:14.3}
	\alpha_i = - \Omega(z,-\mathbb J_{2n}e_i), \quad \beta_i = \Omega(z,e_i),
\end{equation}
for $i=1,\dots,k$. Now we define the modified vectors as
\begin{equation} \label{eq:SyMo:14.4}
	\tilde z = z - \sum_{i=1}^k \Omega(z,-\mathbb J_{2n}e_i) e_i - \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}e_i.
\end{equation}
If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product. Therefore the set span$(\{e_1,\dots,e_{k+1}\})$ forms a Lagrangian subspace and according to theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , -\mathbb J_{2n} e_{k+1} \}$ forms an orthosymplectic basis.

Note that the $SR$ method can be replaced with a backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
}

@@

Suppose that a $2k$ dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors. Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to find the value that maximises the error in the Hamiltonian
\begin{equation} \label{eq:SyMo:14.5}
	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
\end{equation}
The goal is to approximate the Hamiltonian function as well as possible.

We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots 
\begin{equation}
	S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}.
\end{equation} 
The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
{\edit 
\begin{equation} \label{eq:SyMo:14.6}
	z := \underset{s\in S}{\text{argmax }} \| s - A_{2k}{A_{2k}}^+s \|.
\end{equation}
}
Finally, we update the basis as
{\edit
\begin{equation} \label{eq:SyMo:14.7}
	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , -\mathbb J_{2n}e_{k+1} \},
\end{equation}
}
where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.

@@

{\edit
\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Gamma_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,-\mathbb J^T_{2n}e_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| z^*\|$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , -\mathbb J^T_{2n}e_1,\dots,-\mathbb J^T_{2n}e_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}
}

The convergence and convergence rate of the greedy algorithm with respect to Kolmogorov $n$-width is investigated in Section \ref{chap:SyMo.PrSy:3}. We finish this section by stating the advantage of using the greedy algorithm over the POD-based methods:
\begin{itemize}
%\item Lack of a symplectic linear algebra toolbox for solving the minimization in (\ref{eq:SyMo:8}) requires a possibly very expensive nonlinear optimizations to recover near optimal symplectic bases. Alternative approaches for finding solutions to (\ref{eq:SyMo:8}) requires assumptions on the structure of the symplectic basis. The greedy algorithm on the other hand, is a cheap alternative to the nonlinear optimization and it allows flexibility in terms of the structure.
\item The greedy method reduces the total cost of the offline stage by solving the original system (\ref{eq:Hasy:7}) only $k$ times to obtain a reduced system of size $2k$, unlike POD-based methods that require time integration of (\ref{eq:Hasy:7}) for the entire parameter space.
\item If enrichment of the symplectic basis is required, the greedy algorithm can add new basis vectors. The POD-based methods, are bound to re-compute the basis for the refined parameter set.
\item POD-based methods, e.g., the cotangent lift and the complex SVD, require performing SVD decompositions on possibly very large matrices. The greedy algorithm avoids this.
\end{itemize}

@@

	Let $S$ be a compact subset of $\mathbb{R}^{2n}$ with exponentially small Kolmogorov $n$-width $\edit d_{k}\leq c\exp(-\alpha k)$ with $\alpha > \log3$. Then there exists $\beta>0$ such that the symplectic subspaces $A_{2k}$ generated by the greedy algorithm provide exponential approximation properties such that

@@

{\edit
\begin{theorem}
	The basis $A_{2k}$ generated by the greedy algorithm is orthosymplectic.
\end{theorem}
\begin{proof}
	Direct result of theorem \ref{theorem:Hasy:1.6}.
\end{proof}
}
